---
title: "R Markdown for reproducible reports"
author: "David Benkeser, PhD MPH<br> <span style = 'font-size: 50%;'> Emory University<br> Department of Biostatistics and Bioinformatics </span>"
date: 'INFO550<br><br>`r icon::fa("link")` [.white[bit.ly/info550]](https://bit.ly/info550) <br> `r icon::fa("book")` [.white[Additional reading]](https://benkeser.github.io/info550/readings#r-markdown)'
tldr: "Writing reproducible reports has never been easier."
recording: ""
reading: "r-markdown"
output:
  xaringan::moon_reader:
    css: xaringan-themer.css
    nature:
      titleSlideClass: ["left", "middle", "inverse"]
      highlightStyle: tomorrow-night-blue
      highlightLanguage: rmarkdown
      highlightLines: false
---

```{r xaringan-themer, include=FALSE, warning=FALSE}
library(xaringanthemer)

extra_css <- list(
  ".clearbox" = list(
    display = "inline-block", 
    width = "100%",
    padding = "5px",
    border = "1px solid #012169",
    `background-color` = "#d9d9d6",
    `align-items` = "left",
    `vertical-align` = "middle"
  ),
  ".small" =  list(`font-size` = "80%"),
  ".large" =  list(`font-size` = "150%"),
  "ul li" = list(`margin-bottom` = "10px"),
  ".red" = list(color = "#DC143C",
                `font-weight` = "600"),
  ".remark-code-line-highlighted" = list(border = "2px solid #f2a900",
                                         `background-color` = "#002451"),
  ".code-box" = list(display = "inline-block", 
    width = "95%",
    `line-height` = "0.5em",
    padding = "5px",
    `font-size` = "80%",
    `background-color` = "#f6f6f6",
    `align-items` = "left",
    `font-family` = "Courier Prime",
    border = "1px solid black")
)

style_mono_accent(
  base_color = "#012169",
  header_font_google = google_font("DM Sans"),
  text_font_google   = google_font("DM Sans", "400", "400i"),
  code_font_google   = google_font("Courier Prime"),
  extra_css = extra_css
)
```

<style type="text/css">
.remark-slide-content {
    font-size: 25px
}
</style>

### Analysis reports

As a <del>statistician</del> __data scientist__, you must effectively convey results.
* Figures, tables, text describing results, etc...

__Solutions__: 
* email with attachments
* LaTex pdf
* __Rmarkdown__

Why __R markdown__?
* Human readable source
* Highly reproducible reports
* Output html instead of pdf

???

If your job involves a heavy amount of data analysis, you will end up creating
hundreds if not thousands of reports in your career. Invest the time now to 
come up with an efficient system for doing this!

When I was a grad student, there were limited tools for doing this well. sweave
was around, but the learning curve was steep and was not too much easier than 
just using LaTex. The advent of `knitr` and `rmarkdown` was revolutionary for 
making reproducible research more accessible. 

One of the big downsides of LaTex is how ugly the source code is. Want to 
make text bold face `\textbf{something}`. Ditto html. It's clunky. Markdown
has an elegant and simple syntax that is appealing to read.

Reports made with `knitr` can, in theory, be reproduced by anyone, anywhere.
* caveat is you need: same R version, packages, not to use absolute paths, etc..

Recently, I have started using html output for reports over pdf. Appealing
aspects of html include:
* it can be viewed in a web browser (on a mobile device)
* no worrying about page breaks and fitting figures on a page
* heavily customizable, if you know a little css

---

### R Markdown

R Markdown is developed by RStudio
* __R__ : it can incorporate `R` code chunks
	* but also `python`, `bash`, ...
* __Markdown__ : based on markdown syntax
	* but also LaTex, html, ...

No more running code, copy paste results. 
* All automated! 
* Don't touch that mouse!

???

I have my own thoughts about the RStudio GUI, but the best thing the 
RStudio team has done is introduce R Markdown. 

In what follows, code chunks in blue background + highlighting are the code
that might appear in your R Markdown file, i.e., a file with `.Rmd` extension 
that includes your code for building your output document. Objects inside
gray boxes are examples of what the code will look like when rendered. 

Note that if you are viewing the plain text `.Rmd` that I use to make these
slides, the code will look a little more complex, as I have to find ways
of making R Markdown display verbatim code instead of evaluating that code. 

---

### Markdown

```
# This is a section header
## This is a subsection header
### This is a subsubsection header

__This is bold text__. 

*This is italic text*. 

This is `monospace text for inline code`.

~~This text is struckthrough.~~

* A bulleted list
* Another item in my list

1. A numbered list
2. The second item in my list

--- 

text between horizontal rules

---

[Hyperlinks are easy](https://google.com)
```

???

A key design element of Markdown is its readability. The plain text 
files are imminently readable with various extra commands and markup associated
with LaTex or html. 

Many websites allow text input in Markdown (e.g., GitHub comments, reddit). 

---

### R Markdown

```{r, define-my-variable,echo = FALSE}
my_variable <- rnorm(1)
```

R Markdown blends Markdown with code chunks. 

````markdown
Here is some plain text in my document explaining what I did.

`r ''````{r}
# now I want to show some code
my_variable <- rnorm(1)
```
````

This renders as:

.clearbox[
Here is some plain text in my document explaining what I did.

.code-box[
\# now I want to show some code

my_variable <- rnorm(1)
]
]

???

Use three back ticks + a language (in this case, R) to demarcate a "code chunk".
When the document is compiled, code in these code chunks will be run in
a fresh R session, while (before, actually) the document compiles. 

---

### R Markdown

You can also reference variables inline. 

```markdown
The value of the variable I created is `r knitr::inline_expr("my_variable")`.
```

This renders as 

.clearbox[
The value of the variable I created is `r my_variable`.
]

No more copy and pasting results into text!

???


All inline code needs to be on a single line. 
* Helpful to use a hidden code chunk that defines variables and then reference variables by their name in the text. 

---

### yaml header

R Markdown documents begin with a __yaml header__. 
* "meta-data" for the document
* options that control how the document is rendered

A basic yaml header:

````markdown
---
title: "My Great Report"
author: "Brilliant Student"
date: "`r knitr::inline_expr("format(Sys.Date(), '%Y-%m-%d')")`"
output: html_document
---
````

???

The yaml header includes all the information that `knitr` needs to 
know how to compile your document.

Note that in the above, I have included inline R code to be rendered 
when the document compiles.
* `format(Sys.Date(), '%Y-%m-%d')` formats the date in year-month-day format.

---

### yaml header

You can include options in the yaml header. 
* note the nested structure
* don't forget your `:`'s!
* two spaces, __no tabs__!

````markdown
---
title: "My Great Report"
author: "Brilliant Student"
date: "`r knitr::inline_expr("format(Sys.Date(), '%Y-%m-%d')")`"
output: 
  html_document:
    toc: true
    highlights: "pygments"
---
````

???

I am constantly messing up yaml syntax and chasing down mysterious errors. 
Once you have a header you like, I recommend sticking with it unless you have
serious cause to change it. 

There are almost endless options for customizing the looks of documents.

---

### Output formats

There are [many output formats](https://rmarkdown.rstudio.com/lesson-9.html#documents). Here are a few useful ones (with links to 
further documentation): 
.small[
| `output`                                                                                       | Description                                                                                                            |
|------------------------------------------------------------------------------------------------|------------------------------------------------------------------------------------------------------------------------|
| [`html_document`](https://bookdown.org/yihui/rmarkdown/html-document.html)                     | html template with many preset themes                                                                                  |
| [`pdf_document`](https://bookdown.org/yihui/rmarkdown/pdf-document.html)                       | pdf using LaTex                                                                                                        |
| [`word_document`](https://bookdown.org/yihui/rmarkdown/word-document.html)                     | MS Word document (can [create custom theme](https://bookdown.org/yihui/rmarkdown-cookbook/word-template.html)) |
| [`ioslides_presentation`](https://bookdown.org/yihui/rmarkdown/ioslides-presentation.html)     | HTML5 presentation slides                                                                                              |
| [`beamer_presentation`](https://bookdown.org/yihui/rmarkdown/beamer-presentation.html)         | pdf presentation slides with beamer                                                                                    |
| [`powerpoint_presentation`](https://bookdown.org/yihui/rmarkdown/powerpoint-presentation.html) | MS Powerpoint presentation                                                                                             |
]

---

### knitr and pandoc

* `rmarkdown`
	* an `R` package that converts `.Rmd` to different file types
	* an `R` interface to `pandoc`

* `pandoc`
	* a general purpose document converter
	* command line tool, no GUI

* `knitr`
	* an `R` package that executes code chunks
	* "knits" results back to document

???

It's useful to understand which program is doing what in case you 
need to chase down errors. 

Also, note that you need `pandoc` to render R Markdown. If you have R 
Studio installed, `pandoc` comes with it. Otherwise, you need to do a separate
installation. 

---

### Rendering

The "easiest" way to render documents is click "Knit" in RStudio
* Don't touch that mouse!

We will be rendering documents from the command line as follows.


```{bash, render-demo, eval = FALSE}
# from directory where example.Rmd file lives
Rscript -e "rmarkdown::render('example.Rmd')"
```

See the [`render` documentation](https://rmarkdown.rstudio.com/docs/reference/render.html) for options. 

???

Here we are calling `Rscript` (the R executable for scripts) from the command 
line. This starts a new session of `R` and executes the command
`rmarkdown::render('example.Rmd')`, which creates the document `example.html`.

The `::` syntax in the `R` command says, "in the `rmarkdown` package use 
the (exported, in case you care) function `render`. 
* Allows us to avoid loading the library
* e.g., you could also run `Rscript -e "library(rmarkdown); render('example.Rmd')"`, but its more typing. 

---

### Breakout exercise 1

Download <a href="examples/example.Rmd" download>`example.Rmd`</a>. 
* Render the document as is from the command line.
* Without changing `example.Rmd`, render it as an `ioslides_presentation` that is saved as `example_presentation.html`. 
* Add additional options in the yaml header (e.g., change font size, or code highlighting), render the document again and see what changes. 
* In the __Executive Summary__ section, add inline code that reports the number of rows and columns in the `mtcars` data set. Render to make sure it works. 

---

### Code chunk options

* Add short labels explaining what chunks do.
* Include options to control how code chunks behave.

````markdown
Here is some plain text in my document explaining what I did.

`r ''````{r, label-of-chunk, echo = FALSE}
# this code will not appear in my rendered document
# but it is still run
my_variable <- rnorm(1)
```

The value of the variable I created is `r knitr::inline_expr("my_variable")`.
````

This renders as

.clearbox[
Here is some plain text in my document explaining what I did.

The value of the variable I created is `r my_variable`.
]

???

Some notes on chunk labels:
* avoid spaces, underscores, and periods
* hyphens are safe

Options for each code chunk are included, separated by commas. These must all
appear on one line. There are __many__ such options that can be included. 
Here we'll go over just a few. For a full list see 
[here](https://yihui.org/knitr/options/).

---

### Code chunk options

Option values can be any valid `R` expression. For example,

````markdown
Here is some plain text in my document explaining what I did.

`r ''````{r, decide-whether-to-print, echo = FALSE}
# define a variable that determines whether I will print
# a subsequent code chunk
should_i_print <- rnorm(1) > 0
```

`r ''````{r, maybe-gets-printed, echo = should_i_print}
# writing code to print only some of the time
doing_some_more_stuff <- "more cool code"
```
````

---

### Code chunk options

If `should_i_print == TRUE`, this renders as 

.clearbox[
Here is some plain text in my document explaining what I did.

.code-box[
\# writing code to print only some of the time

doing_some_more_stuff <- "more cool code"
]
]


If `should_i_print == FALSE`, this renders as 

.clearbox[
Here is some plain text in my document explaining what I did.
]

---

### Code chunk options


Common options for controlling display of results.

| Option    | Action                                              |
|:----------|:----------------------------------------------------|
| `eval`    | Run the code included in the chunk?                 |
| `echo`    | Show the code chunk in the rendered document?       |
| `warning` | Print warning messages generated by code?           |
| `error`   | Print and keep running after errors?                |
| `message` | Print messages generated by code?                   |
| `include` | Show the code and results in the rendered document? |

---

### Code chunk options

Change default options using code chunk at the top of the `.Rmd`.

````markdown
`r ''````{r, setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, warning = FALSE, fig.width = 6, fig.height = 6
)
```
````

---

### Figures

Figures produced by R code will be placed after the code chunk from which they were generated. 

````markdown
`r ''````{r, make-a-plot}
# make up data
x <- rnorm(10); y <- rnorm(10)
# make a plot
plot(x, y)
```
````

See all [figure options here](https://yihui.org/knitr/options/#plots)

???

The most relevant options are `fig.height`, `fig.width`, and `fig.cap`.

---

### Tables

There are several options for making tables in R Markdown.
* just print a data.frame or matrix
	* + easy, - monospaced format
* [`knitr::kable`](https://bookdown.org/yihui/rmarkdown-cookbook/kable.html)
	* + easy, nice looking, - not 100\% flexible
* [`xtable::xtable`](https://www.rdocumentation.org/packages/xtable/versions/1.8-4/topics/xtable)
	* + highly customizable, - only for pdfs (LaTex based)
* [`pander::pander`](https://cran.r-project.org/web/packages/pander/vignettes/knitr.html)
	* + default tables for different `R` classes
* [`stargazer::stargazer`](https://cran.r-project.org/web/packages/stargazer/vignettes/stargazer.pdf)

???

I find `kable` sufficient for 90\% of my report needs and `xtable` for creating tables for LaTex manuscripts. 

Think about whether there is a need for really nicely formatted tables or whether raw output is OK. Whose the audience?

---

### Referencing

The [`bookdown` package](https://bookdown.org/) allows for table/figure referencing. 
* use e.g., `option: bookdown::html_document2()`

````markdown
As we see in Figure \@ref(fig:make-a-plot), there is no relationship.

`r ''````{r, make-a-plot, echo = FALSE, fig.cap="A great figure I made"}
x <- rnorm(10); y <- rnorm(10)
plot(x, y)
```
````

This renders as 

.clearbox[
As we see in Figure 1, there is no relationship.

[figure here] 
<br> 
.small[Figure 1: A great figure I made]
]

???

A few notes:
* You must have a `fig.cap` for this to work. 
* For tables they must also include a caption (e.g., `caption` option of `kable`)
* If you have a code chunk named `my-chunk` that produces a figure, the reference will be `fig:my-chunk`.
* If you have a code chunk named `my-chunk` that produces a table, the reference will be `tab:my-chunk`.
* Again, avoid underscores in chunk names!

---

### Rounding

Automating rounding can be annoying, but definitely do it!
* `round` drops 0s 
* `sprintf` is better

```{r, round-fn}
good_round <- function(x, digits = 2, pval = FALSE){
	# just use round otherwise
	stopifnot(digits > 1)
	stopifnot(length(digits) == 1)
	if(pval){
		stopifnot(x >= 0 & x <= 1)
	}
	tmp <- sprintf(paste0("%.", digits, "f"), x)
    zero <- paste0("0.", paste0(rep("0", digits), collapse = ""))
    tmp[tmp == paste0("-", zero)] <- zero
    if(pval & tmp == zero){
    	tmp <- paste0("<0.", paste0(rep("0", digits - 1), collapse = ""), "1", 
    	              collapse = "")
    }
    return(tmp)
}
```

???

The `good_round` function solves a couple problems:
* `round` will drop trailing zeros, which makes output inconsistent
* `sprintf` will report e.g., `-0.00`
* both will report p-values of 0, which is generally not good practice

---

### Rounding

```{r show-rounding}
good_round(215.125, digits = 2)
good_round(0.003154, digits = 2)
good_round(0.003154, digits = 2, pval = TRUE)
```

???

Please adhere to rounding conventions in your reports. It's annoying but
your collaborators will thank you. 

---

### Reproducible reports

* .red[Do not use absolute paths in your code!]
	* e.g., `~/my_folder/that_you/dont_have/`
* Keep all code and data in one directory
	* e.g., a GitHub repo
* If you must use absolute paths, define variables with directories at the top of your document and describe fully in `README`.
	* e.g., `save_dir = ~/my_folder`
* For anything with randomness, `set.seed`!
* Include a final chunk that displays the output of `getwd()` and `devtools::session_info()`.
	* information about your system to help others reproduce
* Use `Make` files to help users understand construction
	* next lecture

---

### Breakout exercise 2

Working again off `example.Rmd`:
* add a chunk at the beginning that suppresses all code chunks;
* add inline code in Executive Summary to describe conclusion;
* add code to chunk `make-scatter` that produces a scatter plot and modify options until it looks good;
* add code to chunk `make-table` and modify until it looks good (make sure you round!).

