---
title: "Miles per gallon versus number of cylinders"
author: "INFO550 student"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output: html_document
---

<!--
	Breakout exercise 2:
	Add a code chunk that suppresses all code chunks from appearing
	in the rendered document.
-->

## Executive summary

```{r, load-data}
# load mtcars data
data(mtcars)
```

<!--
	Breakout exercise 1:
	Add inline code below that counts the number of rows and columns
	in each data set. 
-->

There are ... observations in the data set, with ... features measured on each.

<!--
	Breakout exercise 2:
	Insert inline code below that renders as "positive" if the estimated 
	Pearson correlation between mph and cyl is >0 and "negative" if the 
	Pearson correlation is less than or equal to 0. 
-->

Overall, we found a ... relationship between miles per gallon and number of cylinders.

## Analysis

We analyzed the relationship between miles per gallon (`mpg`) and number of 
cylinders (`cyl`). A scatterplot is shown below.

```{r, make-scatter}
# Breakout exercise 2:
# add code to create a scatter plot of mpg ~ cyl
# adjust fig.width, fig.height, other options to make it legible
# in the rendered document
```

A summary of the miles per gallon stratified by cylinders is shown in the Table 
below. 

```{r, make-table}
# get summary statistics by number of cylinders
summary_by_cyl <- by(mtcars, mtcars$cyl, function(x) summary(x$mpg))
stratified_summary <- Reduce(rbind, summary_by_cyl)
row.names(stratified_summary) <- c(4, 6, 8)

# Breakout exercise 2:
# make a table of stratified summary object

```